Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-06-26 papers/paper-eccentricity/paper-eccentricity.tex
Different abstract versions from Overleaf

@@

@@

Firstly, dynamical friction \citep{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scale separations to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{Milosavljevic2001,Merritt2013book}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,Rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep[e.g.][]{Mannerkoski2019,Mannerkoski2022}, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964}.

@@

In this Letter, we find that the SMBH binary eccentricity is reliably predictable only for particular
initial conditions of the galaxy merger system, and does not in general hold.
Indeed, we find that for realistic galaxy merger orbits where the initial orbital eccentricity of the system is $e_0>0.90$, stochasticity in the final
SMBH binary eccentricity is physical and unavoidable.

@@

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{mannerkoski2023,rantala2017}, in which the dynamics of
SMBH and stars in a small region around them are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes Post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{Mora2004}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy is thus represented as a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed.
We test six different mass resolutions %by sampling the stellar phase space distribution
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$.

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
The radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022} for the $e_0=0.90$ orbit.
For each orbital configuration, we run ten realisations for each mass resolution, to account for stochasticity caused by the discretised phase space.

@@

The SMBH binary eccentricity for both the $e_0=0.90$ (blue) and $e_0=0.99$ (orange) orbits are shown in \autoref{fig:eccentricities} as a function of shifted time $t' = t - t_\mathrm{bound}$ for the highest mass resolution sets, where $t_\mathrm{bound}$ is the time when the binary orbital energy $\mathcal{E}$ becomes and remains negative. 
The median time when each simulation set forms a hard binary is similar between the two sets, as indicated by the arrows in \autoref{fig:eccentricities}.
The $e_0=0.90$ set also demonstrates six SMBH-binary mergers within $50\,\mathrm{Myr}$ of forming a bound binary, seen as a rapid orbit circularisation in \autoref{fig:eccentricities}, which is captured self-consistently using \ketju{}.
The $e_0=0.99$ runs show a wide variation in eccentricity, where $e$ spans almost the entire domain range $[0,1]$.
Even though the $e_0=0.99$ runs have a higher initial merger eccentricity than the $e_0=0.90$ runs, none of the SMBH binaries in the shown $e_0=0.99$ set obtain high enough eccentricities to undergo GW-induced coalescence.

@@

We show the dependence of inter-simulation eccentricity standard deviation on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with circle markers, and for the $e_0=0.99$ orbit using square markers.

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger, in agreement with the results of \citet{nasim2020}. 
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.99$ orbit. 
For mass resolutions $M_\bullet/m_\star<10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions. \drafting{Add about 4M runs when done}

@@

\begin{figure*}
    \centering
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to a time shortly after $t_\mathrm{bound}$ of a single SMBH in two representative realisations of the $e_0=0.90$ mergers (left) and the $e_0=0.99$ mergers (right), which by symmetry of the equal mass system is a reflection of the second BH orbit about the $x=-z$ line. 
    Each line is coloured according to the shifted time $t'=t-t_\mathrm{bound}$ of the simulation, however the colours for $|t'|>2\,\mathrm{Myr}$ is constant for visual clarity. % of the deflection-phase of the evolution. 
    In each inset panel, $\theta_\mathrm{defl}$ is the deflection between the arrow heads.
    The circle marker indicates the SMBH position just before a bound binary is formed.}
    \label{fig:orbit}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Scattering deflection angle $\theta_\mathrm{defl}$ and the resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to reasonably well fitting model curves.
    The solid lines show the results from the toy model with $e_\mathrm{s}=0.91$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.90$ and $e_\mathrm{s}=0.92$.
    The model curves have been shifted left by $12\degr$ and $6\degr$
    to better match the simulation data, with the shift also indicated by arrows.
    The marginal histograms show kernel density estimates to the simulation data.
    }
    \label{fig:theta_e}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics{sample_model_orbits}
    \caption{
    Top:
    Sample orbits computed with the toy model for different background potential eccentricities $e_\mathrm{s}$,
    with otherwise identical initial conditions.
    The dashed lines show the isopotential contours of the model.
    Only the early part of the evolution is shown for clarity.
    Bottom:
    The orbital eccentricity $e$ of the model orbits (note the non-linear scale).
    The vertical line marks the end of the period shown in the top panels. 
    }
    \label{fig:model_orbits}
\end{figure*}

The dependency of the binary eccentricity on the 
deflection angle $\theta_\mathrm{defl}$ during the binary formation
can be reproduced using a simple toy model,
which includes only the essential components relevant to this process.
Taking the orbit to lie in the $xy$-plane,
the relative motion of the two equal-mass SMBHs in this model follows the 
equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the SMBHs,
$M_\bullet = 10^8\,\Msun$ is the mass of a single SMBH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the asymmetric background potential,
and $\vb{a}_\mathrm{DF}$ the acceleration due to dynamical friction.

The stellar background can be modeled as a constant density spheroidal potential
\citep[e.g.][]{binney2008}
\begin{equation}
\Phi_\mathrm{bg}(\vb{x}) = \pi G \rho (A_x x^2 + A_y y^2),
\end{equation}
with the acceleration given by 
\begin{equation}
\vb{a}_\mathrm{bg}(\vb{x}) = - \nabla \Phi_\mathrm{bg}(\vb{x}).
\end{equation}
The $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{align}
A_x &= 2 \left(\frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2}\right) 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]\\
A_y &= \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right],
\end{align}
with the long axis aligned along the $x$-axis.
The stellar density is set to $\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
which approximately matches the values seen in the simulations when the SMBH binary is becoming bound.

The dynamical friction of a single BH is modeled using the \citet{chandrasekhar1943} formula assuming
a Maxwellian distribution with a constant velocity dispersion $\sigma_\star$ \citep[e.g.][]{binney2008}:
\begin{equation}\label{eq:chandra_df}
\begin{aligned}
\vb{a}_{1,\mathrm{DF}} = &-\frac{4 \pi G^2 M_\bullet \rho \ln{\Lambda}}{|\vb{v}_{1}|^3}\\
                         &\times\left(\operatorname{erf}(X) - \frac{2X}{\sqrt{\pi}} \exp(-X^2)\right)\vb{v}_{1},
\end{aligned}
\end{equation}
where $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH,
and $X = |\vb{v}_1|/\sqrt{2}\sigma_\star$.
We set ${\sigma_\star=200\,\mathrm{km\,s^{-1}}}$ based on the value measured from the simulations.
The value of the Coulomb logarithm is expected to be $\ln{\Lambda}\sim 4\textnormal{--}5$
based on the size of the stellar system,
and we find $\ln{\Lambda} = 4.7$ to give results that agree well with the simulations.

The effect of dynamical friction weakens as the SMBH binary orbit shrinks.
To account for this, when the binary is bound 
we multiply the acceleration given by Eq.~\eqref{eq:chandra_df}
with the smooth cut-off function
\begin{equation}
f(a) = \frac{1}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]}.
\end{equation}
Here $a$ is the semi-major axis of a bound binary,
and the cut-off scales are $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ defined by equation \eqref{eq:ahard}.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

To mimic the nearly linearly plunging orbits before the scattering event seen in \autoref{fig:orbit},
we specify the initial conditions as 
\begin{align}
\vb{x}_0 &= (25\,\mathrm{pc}, b)\\
\dot{\vb{x}}_0 &= (-v_0, 0).
\end{align}
In order to match the typical SMBH relative velocity seen at this separation in
the simulations we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ for the $e_0=0.90$ case, and
 $v_0 = 560\,\mathrm{km\,s^{-1}}$ for the the $e_0=0.99$ case.
The spheroid eccentricity parameter $e_\mathrm{s}$ is not easily measured from the simulations,
due to stellar components that remain tightly bound to the SMBHs and obscure the background
potential relevant for the dynamics,
as well as due to fact that the potential in the simulations is constantly evolving.
However, only sufficiently large values of $e_\mathrm{s}$ allow for low binary
eccentricities to be produced,
as is shown by \autoref{fig:model_orbits}.
Performing calculations with different values of $e_\mathrm{s}$,
we found a good fit to the simulation results for $e_\mathrm{s}\approx 0.9$.

@@

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
These shifts are required due to the rotation of the stellar component tilting the background potential
relative to the SMBH trajectories in the simulations, with the effect being stronger in the $e_0=0.90$ case
due to the less radial merger orbit.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the model does appear to capture the mean behaviour fairly well, apart from the largest values of
$\theta_\mathrm{defl}$.
The behaviour of the model in this region is also quite sensitive to the values of $e_\mathrm{s}$
and $\ln{\Lambda}$,
which might be related to the large scatter in the simulation data.

Otherwise the behaviour of the simulation data is captured well,
with the model correctly producing the two main eccentricity minima,
as well as the $e\approx 1$ region between them.
The minima occur for trajectories where the torque from the background potential together with the dynamical friction
causes the SMBHs to loop around into a nearly circular orbit,
while highly eccentric binaries are produced when the binary is trapped into nearly radial oscillations along 
the main axes of the potential, as is seen in \autoref{fig:model_orbits}.

@@

Our results show that the variation in eccentricity $e_\mathrm{h}$ for a given system configuration is tightly correlated with the variation in the deflection angle $\theta_\mathrm{defl}$, independent of the simulation mass resolution.
Together with the simple toy model that reproduces well the behaviour seen in the simulations,
this demonstrates that the sensitivity of the SMBH binary eccentricity to slight changes in the merger orbit is a physical feature of galactic mergers on highly radial orbits,
at least in major mergers where the SMBHs have nearly equal masses,
and not an error due to insufficient numerical resolution.

As shown by both the simulations and the toy model, in these mergers the eccentricity of the SMBH binary can span nearly the full possible range between 0 and 1,
depending on comparatively tiny differences on the scale of a parsec in the particular realisation of the galactic merger orbit.
Compared to major mergers, in minor mergers where there is less asymmetry in the stellar background when the SMBHs become bound,
the scatter in eccentricity is likely to be low even in the case of a radial merger orbit, since the system is less sensitive to the exact value of the deflection angle,
or the impact parameter, as can be seen from the $e_\mathrm{s}=0.2$ case in \autoref{fig:model_orbits}.

The simulation setup used in this work is highly idealised,
and covers only a small part of the parameter space of merger configurations since we only consider fairly high-eccentricity galactic merger orbits.
However, similar behaviour can be expected to occur also in more realistic models,
since the relevant dynamics occurs within a few hundred parsecs from the centre of the merged galaxy,
where we do not expect any significant  effects from dark matter or the outer parts of the galaxy, which are not included in our simulations.
Lower eccentricity merger orbits can be expected to show less scatter in the binary eccentricity due to the lack of a hard scattering event that is sensitive to slight perturbations in the merger orbit,
and this is also seen in simulation results \citep{gualandris2022}.
However, most major mergers are expected to occur on highly radial orbits based on cosmological simulations \citep[e.g.][]{khochfar2006},
and the dynamical friction from the dark matter halo would generally drive even lower eccentricity merger orbits to highly radial final plunges when the initial separation of the galaxies is large enough.

>>>>

We study the stochasticity of supermassive black hole (SMBH) binary eccentricity in $N$-body simulations of equal-mass galaxy mergers using the \ketju{} code.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and the binary eccentricity is captured using a simple toy model, indicating that it is driven by the interplay between an asymmetric stellar potential and dynamical friction acting on the SMBHs, with no apparent resolution dependence.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in binary eccentricities varying in nearly the full possible range between 0 and 1.
By increasing the number of stellar particles, the reduction in Brownian motion means a smaller range of deflection angles is sampled, thus allowing the binary eccentricity produced by a given merger orbit to converge towards a definite value.
We conclude that the sensitivity of the eccentricity to small-scale perturbations of the merger orbit means that the binary eccentricity is effectively a random quantity when considering realistic galaxy mergers, however we stress that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.

@@

\drafting{Another version:}

\drafting{
We study the stochasticity of the eccentricity of SMBH binaries formed in $N$-body simulations equal-mass galactic mergers.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and binary eccentricity is reproduced with a simple toy model, showing that it is caused by an interplay between the asymmetric stellar potential and dynamical friction acting on the SMBHs.
There is also no apparent resolution dependency in this mapping, with lower particle count simulations simply sampling a wider range of deflection angles due to Brownian motion.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in eccentricities varying in nearly the full possible range between 0 and 1.
Higher particle counts reduce the Brownian motion induced variation in the merger orbit in idealised simulations, allowing the eccentricity produced by a given merger orbit to converge towards a definite value.
However, we argue that the sensitivity of the eccentricity on the small-scale perturbations of the merger orbit means that the binary eccentricity is an effectively random quantity when considering realistic galaxy mergers, but also that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.
}

\drafting{Third version:}

\drafting{
We study the stochasticity of the eccentricity of SMBH binaries formed in $N$-body simulations of equal-mass galactic mergers, run with the \ketju{} code. In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the final hard scattering, as defined by the deflection angle. The 
mapping between the deflection angle and the binary eccentricity can be reproduced with a simple toy model, showing that it is caused by an interplay between the asymmetric stellar potential and dynamical friction acting on the SMBHs, with no apparent resolution dependency.
At lower particle resolutions the increase in Brownian motion result in a wider sampling of the deflection angles, whereas at higher particle counts a given merger orbit converge towards a more definite eccentricity value.
Due to the non-linearity of this mapping, for certain merger configurations small, parsec-scale variations in the merger orbit can result in final SMBH eccentricities that nearly cover the full $e=[0,1]$ eccentricity range. 
We conclude that the sensitivity of the eccentricity on the small-scale perturbations of the merger orbit means that the binary eccentricity is effectively a random quantity when considering realistic galaxy mergers, however we stress that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.
}

@@

Firstly, dynamical friction \citep{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{Milosavljevic2001,Merritt2013book}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,Rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep[e.g.][]{Mannerkoski2019,Mannerkoski2022}, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964}.

@@

In this Letter, we find that the SMBH binary eccentricity is reliably predictable only for particular
initial conditions of the galaxy merger system, and does not in general hold.
Indeed, we find that for realistic galaxy merger orbits where the SMBHs undergo a final near-radial plunge before becoming bound, stochasticity in the final
%initial orbital eccentricity of the system is $e_0>0.90$, stochasticity in the final
SMBH binary eccentricity is physical and unavoidable.

@@

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{mannerkoski2023,rantala2017}, in which the dynamics of
SMBH and stars in a small region around them are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes Post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{Blanchet2014}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy consists of a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed.
We test six different mass resolutions 
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$, corresponding to $M_\bullet/m_\star= 1000 \textnormal{--} 40000$. \drafting{See if we include even higher res runs.}

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
The radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022} for the $e_0=0.90$ orbit.
For each orbital configuration, we run ten realisations for each mass resolution, to account for stochasticity caused by the discretised phase space. 
Interactions between stellar particles are softened with a softening length of $\epsilon = 2.5\,\mathrm{pc}$.

@@

The SMBH binary eccentricity for both the $e_0=0.90$ (blue) and $e_0=0.99$ (orange) orbits are shown in \autoref{fig:eccentricities} as a function of shifted time $t' = t - t_\mathrm{bound}$, where $t_\mathrm{bound}$ is the time when the binary orbital energy $E$ becomes permanently negative. 
The median time when each simulation set forms a hard binary is similar between the two sets, as indicated by the arrows in \autoref{fig:eccentricities}.
The $e_0=0.90$ set also demonstrates six SMBH-binary mergers within $50\,\mathrm{Myr}$ of forming a bound binary, seen as a rapid orbit circularisation in \autoref{fig:eccentricities}, which is captured self-consistently using \ketju{}.
The $e_0=0.99$ runs show a wide variation in eccentricity, where $e$ spans almost the entire domain range $[0,1]$.
Even though the $e_0=0.99$ runs have a higher initial merger eccentricity than the $e_0=0.90$ runs, none of the SMBH binaries in the shown $e_0=0.99$ set obtain high enough eccentricities to undergo GW-induced coalescence during the $50\,\mathrm{Myr}$ timespan.

@@

We show the dependence of the inter-simulation eccentricity standard deviation on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with blue circle markers, and for the $e_0=0.99$ orbit using orange square markers.

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger, in agreement with the results of \citet{nasim2020}. 
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.90$ orbit. 
For mass resolutions $M_\bullet/m_\star<10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions. \drafting{Add about 4M runs when done}

@@

\begin{figure*}
    \centering
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to a time shortly after $t_\mathrm{bound}$ of a single SMBH in two representative realisations of the $e_0=0.90$ mergers (left) and the $e_0=0.99$ mergers (right), which by symmetry of the equal mass system is a reflection of the second BH orbit about the $x=-z$ line. 
    Each line is coloured according to the shifted time $t'=t-t_\mathrm{bound}$ of the simulation, however the colours for $|t'|>2\,\mathrm{Myr}$ is constant for visual clarity. % of the deflection-phase of the evolution. 
    In each inset panel, $\theta_\mathrm{defl}$ is the deflection at the pericentre between the arrows.
    The circle marker indicates the SMBH position just before a bound binary is formed.}
    \label{fig:orbit}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Scattering deflection angle $\theta_\mathrm{defl}$ and the resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to fitted model curves.
    The solid lines show the results from the toy model with $e_\mathrm{s}=0.91$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.90$ and $e_\mathrm{s}=0.92$.
    The $e_0=0.90$ model curves have been shifted left by $14\degr$
    to better match the simulation data, with the shift also indicated by an arrow.
    The marginal histograms show kernel density estimates of the simulation data.
    The inset panel for the $e_0=0.90$ data shows a zoom-in with $\theta_\mathrm{defl}=[52\degr, 73\degr]$ and $e_\mathrm{h}=[0.96, 1.00]$ for clarity.
    }
    \label{fig:theta_e}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics{sample_model_orbits}
    \caption{
    Top:
    Sample orbits computed with the toy model for different background potential eccentricities $e_\mathrm{s}$,
    with otherwise identical initial conditions.
    The dashed lines show the isopotential contours of the model.
    Only the early part of the orbital evolution is shown for clarity.
    Bottom:
    The orbital eccentricity $e$ of the model orbits (note the non-linear scale).
    The vertical line marks the end of the period shown in the top panels. 
    }
    \label{fig:model_orbits}
\end{figure*}

The dependency of the binary eccentricity on the 
deflection angle $\theta_\mathrm{defl}$ during the binary formation
can be reproduced using a simple toy model,
which includes only the essential components relevant to this process.
Taking the orbit to lie in the $xy$-plane,
the relative motion of the two equal-mass SMBHs in this model follows the 
equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the SMBHs,
$M_\bullet = 10^8\,\Msun$ is the mass of a single SMBH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the asymmetric background potential,
and $\vb{a}_\mathrm{DF}$ is the acceleration due to dynamical friction.

The stellar background is modeled as a constant density spheroidal potential
\citep[e.g.][]{binney2008}
\begin{equation}
\Phi_\mathrm{bg}(\vb{x}) = \pi G \rho (A_x x^2 + A_y y^2),
\end{equation}
with the acceleration given by 
\begin{equation}
\vb{a}_\mathrm{bg}(\vb{x}) = - \nabla \Phi_\mathrm{bg}(\vb{x}).
\end{equation}
The $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{align}
A_x &= 2 \left(\frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2}\right) 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]\\
A_y &= \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right],
\end{align}
with the long axis aligned along the $x$-axis.
The stellar density is set to $\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
which approximately matches the values seen in the simulations when the SMBH binary is becoming bound.

The dynamical friction acting on a single BH is modeled using the \citet{chandrasekhar1943} formula assuming
a Maxwellian distribution with a constant velocity dispersion $\sigma_\star$ \citep[e.g.][]{binney2008}:
\begin{equation}\label{eq:chandra_df}
\begin{aligned}
\vb{a}_{1,\mathrm{DF}} = &-\frac{4 \pi G^2 M_\bullet \rho \ln{\Lambda}}{|\vb{v}_{1}|^3}\\
                         &\times\left(\operatorname{erf}(X) - \frac{2X}{\sqrt{\pi}} \exp(-X^2)\right)\vb{v}_{1},
\end{aligned}
\end{equation}
where $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH,
and $X = |\vb{v}_1|/\sqrt{2}\sigma_\star$.
We set ${\sigma_\star=200\,\mathrm{km\,s^{-1}}}$ based on the value measured from the simulations.
The value of the Coulomb logarithm is expected to be $\ln{\Lambda}\sim 4\textnormal{--}5$
based on the size of the stellar system,
and we find that $\ln{\Lambda} = 4.7$ gives results that agree well with the simulations.

The effect of dynamical friction weakens as the SMBH binary orbit shrinks.
To account for this, when the binary is bound 
we multiply the acceleration given by equation~\eqref{eq:chandra_df}
with the smooth cut-off function
\begin{equation}
f(a) = \frac{1}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]}.
\end{equation}
Here $a$ is the semi-major axis of a bound binary,
and the cut-off scales are $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ defined by equation~\eqref{eq:ahard}.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

To mimic the nearly linearly plunging orbits before the scattering event seen in \autoref{fig:orbit},
we specify the initial conditions as 
\begin{align}
\vb{x}_0 &= (25\,\mathrm{pc}, b)\\
\dot{\vb{x}}_0 &= (-v_0, 0).
\end{align}
In order to match the typical SMBH relative velocity seen at this separation in
the simulations we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ for the $e_0=0.90$ case, and
 $v_0 = 560\,\mathrm{km\,s^{-1}}$ for the the $e_0=0.99$ case.

The spheroid eccentricity parameter $e_\mathrm{s}$ is not easily measured from the simulations,
due to stellar components that remain tightly bound to the SMBHs and obscure the background
potential relevant for the dynamics,
as well as due to fact that the potential in the simulations is constantly evolving.
However, only sufficiently large values of $e_\mathrm{s}$ allow for low binary
eccentricities to be produced,
as is shown by \autoref{fig:model_orbits}.
Performing calculations with different values of $e_\mathrm{s}$,
we found a good fit to the simulation results for $e_\mathrm{s}\approx 0.9$.

@@

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $14\degr$.
This shift is required to account for the rotation of the stellar component tilting the background potential
relative to the SMBH trajectories in the simulations, with the effect being only evident in the $e_0=0.90$ case
due to the less radial merger orbit.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the simple model does appear to capture the mean behaviour relatively well, apart from the largest values of
$\theta_\mathrm{defl}$.
The behaviour of the model in this region is also quite sensitive to the values of $e_\mathrm{s}$
and $\ln{\Lambda}$,
which might be related to the large scatter seen in the simulation data.

However, in general the behaviour of the simulation data is captured well,
with the model correctly producing the two main eccentricity minima,
as well as the $e\approx 1$ region between them.
The minima occur for trajectories where the torque from the background potential together with the dynamical friction
causes the SMBHs to loop around into a nearly circular orbit,
while highly eccentric binaries are produced when the binary is trapped into nearly radial oscillations along 
the main axes of the potential, as is seen in \autoref{fig:model_orbits}.

@@

We find that the variation in eccentricity $e_\mathrm{h}$ for a given system configuration is tightly correlated with the variation in the deflection angle $\theta_\mathrm{defl}$, independent of the simulation mass resolution.
Together with the simple toy model that reproduces well the behaviour seen in the simulations,
this demonstrates that the sensitivity of the SMBH binary eccentricity to slight changes in the merger orbit is a physical feature of galactic major mergers on highly radial orbits, and not an error due to insufficient numerical resolution.
%at least in major mergers where the SMBHs have nearly equal masses,
%and not an error due to insufficient numerical resolution.

As shown by both the simulations and the toy model, the eccentricity of the SMBH binary can span nearly the full possible range between $e=[0,1]$, depending on comparatively small differences on the scale of a few parsecs in the particular realisation of the galactic merger orbit. On the other hand in minor mergers where the stellar background is less asymmetric when the SMBHs become bound,
the scatter in eccentricity is likely to be relatively low even in the case of a radial merger orbit, since the system is less sensitive to the exact value of the deflection angle,
as can be seen from the $e_\mathrm{s}=0.2$ case in \autoref{fig:model_orbits}.

Lower eccentricity merger orbits can be expected to show less scatter in the binary eccentricity due to the lack of a hard scattering event that is sensitive to slight perturbations in the merger orbit, which was also seen in the study by \citet{gualandris2022}.
However, major mergers are expected to occur on highly radial orbits based on cosmological simulations \citep[e.g.][]{khochfar2006}. In addition, the dynamical friction from the dark matter halo would in general drive even lower eccentricity merger orbits to highly radial final plunges when the initial separation of the galaxies is large enough.

The simulation setup used in this work is idealised,
and covers only a small part of the parameter space of merger configurations since we only consider relatively high-eccentricity galactic merger orbits.
However, similar behaviour can be expected to occur also in more realistic models,
since the relevant dynamics occurs within a few hundred parsecs from the centre of the merged galaxy,
where we do not expect any significant effects from dark matter or the outer parts of the galaxy, which are not included in our present simulations.
